\begin{cnabstract}{脂肪酸甲酯；H-ZSM-5分子筛；分子模拟；吸附；扩散}
\par{本文应用Materials Studio 8.0分子模拟软件在623K$\sim$773K，0$\sim$500kPa下研究了脂肪酸甲酯在微介孔H-ZSM-5分子筛中的吸附和扩散行为，考察了孔径、温度、不饱和度和链长对于吸附扩散性能的影响，计算得到了脂肪酸甲酯在分子筛上的吸附等温线、吸附热、吸附位以及扩散系数。吸附结果表明：丁酸甲酯和丁烯酸甲酯在微介孔分子筛内的吸附等温线均满足朗格缪尔吸附模型，在单位质量下不同孔径的分子筛内的饱和吸附量顺序为：20Å介孔 > 60Å介孔 > 微孔；丁烯酸甲酯的吸附量要高于丁酸甲酯的吸附量；在微孔内的等量吸附热随压力升高逐渐增大，而介孔内的等量吸附热随压力升高呈现先增大后减小的趋势；丁酸甲酯优先吸附在交叉孔道，随着压力的增大，直孔道内也有吸附。扩散结果表明：丁酸甲酯在不同孔径下的扩散系数大小顺序为：60Å介孔 ＞ 20Å介孔 $\gg$ 微孔。}
\end{cnabstract}

\begin{enabstract}{fatty acid methyl ester; H-ZSM-5 zeolite; molecular simulation; adsorption; diffusion}
\par{Materials Studio 8.0 molecular simulation software was used to study the adsorption and diffusion behavior of fatty acid methyl ester in micro-mesoporous H-ZSM-5 zeolite at 623K$\sim$\\773K and 0$\sim$500kPa. The effects of pore size, temperature, unsaturation and chain length on the adsorption and diffusion properties were investigated. The adsorption isotherms, adsorption heat, adsorption sites and diffusion coefficients of fatty acid methyl ester on the zeolite were calculated. The results show that: the adsorption isotherms of methyl butyrate and methyl butyrate in micro-mesoporous zeolite satisfy Langmuir adsorption model; the order of saturated adsorption capacity in zeolite with different pore sizes per unit mass is: 20Å mesoporous > 60Å mesoporous > microporous; the adsorption capacity of methyl butyrate is higher than that of methyl butyrate; the isosteric heat of adsorption in the micropore increases with the increase of pressure, while the isosteric heat of adsorption in the mesoporous increases first and then decreases with the increase of pressure; methyl butyrate is preferentially adsorbed in the cross-channel, with the increase of pressure, it is also adsorbed in the straight-channel. Diffusion results show that: the order of diffusion coefficients of methyl butyrate under different pore sizes is: 60Å mesoporous > 20Å mesoporous $\gg$ microporous.}
\end{enabstract}
